---
title: "CS 372 EDA"
author: "Usman Khaliq"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(sf)
library(dcl)
library(tidycensus)
library(lubridate)
library(readxl)
# Parameters
  # US Census Bureau API query for Vintage 2018 County Population Estimates
us_census_vintage_2018 <-
  "https://api.census.gov/data/2018/pep/population?get=GEO_ID,GEONAME,DATE_CODE,DATE_DESC,POP&for=county:*"  


# URL for downloading COVID19 County data from New York Times
covid19_county_nyt <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/live/us-counties.csv"

# Data File for US Census Income and Benefits File 2018
income_and_benefits <- "/Users/usmankhaliq/Downloads/subset2018IncomeAndBenefits.csv" 

#Data file that contains hospital beds in each county
hospital_beds <- "/Users/usmankhaliq/Downloads/subset_HospitalBedData.xlsx"

#Data file that contains demographic data
demographic_data <- "/Users/usmankhaliq/Downloads/demographicData2018_RAEdits.xlsx" 

#Data file that contains data about areas of counties 
counties_area <- "/Users/usmankhaliq/Downloads/LND01.xls"

  # Map colors
MAP_COLORS <- RColorBrewer::brewer.pal(n = 9, name = "PuRd")
MAP_NA_COLOR <- "grey75" 

#Colors for the income and benefits chloropleth maps made
orange_pal <- RColorBrewer::brewer.pal(n = 6, name = "Oranges")

#Colors for hospital beds
blue_pals <- RColorBrewer::brewer.pal(n = 9, name = "Blues") 

#Colors for proportion of blacks
green_pal <- RColorBrewer::brewer.pal(n = 9, name = "Greens")  

#Colors for the population density chart
yellow_pal <- RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")

#state codes for states on east coast
# new york, new jersey, conneticut, massachusetts, california
east_coast_states <- c("36", "34", "25", "09", "06")

#state codes for states in south
south_states <- c("22", "05", "28", "37", "45", "47", "01", "12") 
#===============================================================================

# Code
```


### Downloadig Data

Lets download the most recent data of the current rates of COVID-19 from the New York Times

```{r}
covid19_county_df <- 
  covid19_county_nyt %>% 
  read.csv() %>% 
  rename_all(str_to_lower) %>% 
  mutate(fips = as.integer(fips))
```


Lets download data about populations from counties for 2018.

```{r}
county_populations <-
us_census_vintage_2018 %>%
  jsonlite::fromJSON() %>%
  as_tibble() %>%
  janitor::row_to_names(row_number = 1) %>%
  filter(str_detect(DATE_DESC, "^7/1/\\d{4}\\spopulation\\sestimate")) %>%
  separate(GEO_ID, c("redundant_value", "fips"), sep = "US") %>%
  mutate(
    year = str_extract(DATE_DESC, pattern = "(?<=\\d/\\d/)\\d*(?=\\s.*)")
  ) %>% 
  filter(year %in% c("2018")) %>% 
  select(
    name = GEONAME,
    fips = fips,
    year = year,
    population = POP
  ) %>% 
  mutate(
    fips = as.integer(fips),
    year = as.integer(year),
    population = as.double(population)
  ) %>% 
  arrange(fips, year)
```

Lets read the boundaries of US Counties

```{r}
county_boundaries <- 
  ussf::boundaries(
    geography = "county",
    resolution = "20m"
  ) %>% 
  mutate(fips = as.integer(GEOID)) 
```

```{r}
state_boundaries <- 
  ussf::boundaries(
    geography = "state",
    resolution = "20m"
  )
```



Lets join the COVID-19 county data with the county boundaries.

plot a simple chloropleth map of the county population densities.

```{r}
counties_data <- 
  county_boundaries %>% 
  mutate(GEOID = as.integer(GEOID)) %>% 
  left_join(county_populations, by = c("GEOID" = "fips")) %>% 
  left_join(covid19_county_df, by = c("GEOID" = "fips")) %>% 
  mutate(
    cases_per_100000 = (cases / population) * 100000,
    deaths_per_100000 = (deaths / population) * 100000
  )
```

```{r}
quantile(counties_data$cases_per_100000, na.rm = TRUE)
```


```{r}

labels <- function(x) {
  if_else(x < 379, as.character(x), "140+ ")
}
counties_data %>% 
  mutate(cases_per_100000 = pmin(379, cases_per_100000)) %>% 
  ggplot() +
  geom_sf(aes(fill = cases_per_100000), size = 0.01) +
  geom_sf(data = state_boundaries, color = "grey25", fill = NA, size = 0.2) +
  scale_fill_gradientn(
    #limits = c(0, 150),
    #breaks = seq(0, 150, 75),
    labels = labels,
    colors = MAP_COLORS,
    na.value = MAP_NA_COLOR
  ) +
  guides(
    fill = 
      guide_colorbar(
        barheight = 0.5,
        barwidth = 15,
        title = NULL
      )
  ) +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.5, 0.95),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Number of COVID-19 Cases Per 100,000 People",
    caption = "Sources: US Census Bureau \nNew York Times\nAs of June 9, 2020"
  )
```

```{r}
quantile(counties_data$deaths_per_100000, na.rm = TRUE)

```



```{r}

#quantile(counties_data$deaths_per_100000, na.rm = TRUE)

counties_data %>% 
  mutate(deaths_per_100000 = pmin(13, deaths_per_100000)) %>% 
  ggplot() +
  geom_sf(aes(fill = deaths_per_100000), size = 0.01) +
  geom_sf(data = state_boundaries, color = "grey25", fill = NA, size = 0.2) +
  scale_fill_gradientn(
    #limits = c(0, 150),
    #breaks = seq(0, 150, 75),
    labels = labels,
    colors = MAP_COLORS,
    na.value = MAP_NA_COLOR
  ) +
  guides(
    fill = 
      guide_colorbar(
        barheight = 0.5,
        barwidth = 15,
        title = NULL
      )
  ) +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.5, 0.95),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Number of COVID-19 Deaths Per 100,000 People",
    caption = "Sources: US Census Bureau \nNew York Times\nAs of June 9, 2020"
  )

```



```{r}
quantile(counties_data$cases_per_100000, na.rm = TRUE)
```

```{r}
quantile(counties_data$deaths_per_100000, na.rm = TRUE)
```



### Plotting Chloroleth maps for Food Stamp beneficiaries in Counties

Lets load the data for the foodstamp beneficiaries. We will also extract the FIPS from the GEO_ID column so that it can be combined with the county population values. 

```{r}
income_and_benefits_df <-
  income_and_benefits %>% 
  read_csv() %>% 
  mutate(geo_id = as.integer(str_remove(GEO_ID, "\\d*US")))
```

Lets join the `income_and_benefits_df` with the `county_boundaries` dataframe.

```{r}
counties_income_and_benefit_df <- 
  county_boundaries %>% 
  mutate(GEOID = as.integer(GEOID)) %>% 
  left_join(income_and_benefits_df, by = c("GEOID" = "geo_id"))

```



Now, lets plot a chloropleth map of the percentage of total households in counties on food stamps. 

```{r}
quantile(counties_income_and_benefit_df$Percent_total_house_social_sec_inc, na.rm = TRUE)
```


```{r}
counties_income_and_benefit_df %>% 
  mutate(
    percent_total_house_social_sec_inc = 
      pmin(7.357510, Percent_total_house_social_sec_inc)
  ) %>% 
  ggplot() +
  geom_sf(aes(fill = percent_total_house_social_sec_inc), size = 0.01) +
  geom_sf(data = state_boundaries, color = "grey25", fill = NA, size = 0.2) +
  scale_fill_gradientn(
    #limits = c(0, 150),
    #breaks = seq(0, 150, 75),
    labels = labels,
    colors = orange_pal,
    na.value = MAP_NA_COLOR
  ) +
  guides(
    fill = 
      guide_colorbar(
        barheight = 0.5,
        barwidth = 15,
        title = NULL
      )
  ) +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.5, 0.95),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Percentage of Households on Social Security Income",
    caption = "Source: US Census Bureau Income and Benefits in Counties 2018"
  )

```

Now, lets plot a chloropleth map for percentage of households on cash benefits.

```{r}
quantile(counties_income_and_benefit_df$Percent_total_house_cash_ben, na.rm = TRUE)

```


```{r}
counties_income_and_benefit_df %>% 
  mutate(
    percent_total_house_cash_ben = 
      pmin(2.873391, Percent_total_house_cash_ben)
  ) %>% 
  ggplot() +
  geom_sf(aes(fill = percent_total_house_cash_ben), size = 0.01) +
  geom_sf(data = state_boundaries, color = "grey25", fill = NA, size = 0.2) +
  scale_fill_gradientn(
    #limits = c(0, 150),
    #breaks = seq(0, 150, 75),
    labels = labels,
    colors = orange_pal,
    na.value = MAP_NA_COLOR
  ) +
  guides(
    fill = 
      guide_colorbar(
        barheight = 0.5,
        barwidth = 15,
        title = NULL
      )
  ) +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.5, 0.95),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Percentage of Households on Cash Benefits",
    caption = "Source: US Census Bureau Income and Benefits in Counties 2018"
  )

```

Now, lets plot the Proportion of Households that received food stamps.

```{r}
quantile(counties_income_and_benefit_df$Percent_total_house_food_ben, na.rm = TRUE)

```

```{r}
counties_income_and_benefit_df %>% 
  mutate(
    percent_total_house_food_ben = 
      pmin(16.748792, Percent_total_house_food_ben)
  ) %>% 
  ggplot() +
  geom_sf(aes(fill = percent_total_house_food_ben), size = 0.01) +
  geom_sf(data = state_boundaries, color = "grey25", fill = NA, size = 0.2) +
  scale_fill_gradientn(
    #limits = c(0, 150),
    #breaks = seq(0, 150, 75),
    labels = labels,
    colors = orange_pal,
    na.value = MAP_NA_COLOR
  ) +
  guides(
    fill = 
      guide_colorbar(
        barheight = 0.5,
        barwidth = 15,
        title = NULL
      )
  ) +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.5, 0.95),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Percentage of Households on Food Benefits",
    caption = "Source: US Census Bureau Income and Benefits in Counties 2018"
  )

```


Lets plot the number of beds available in different counties .

```{r}
hospital_data_df <-
  hospital_beds %>% 
  read_excel() %>% 
  mutate(FIPS = as.integer(FIPS))
```

```{r}
hospital_beds_df <- 
  county_boundaries %>% 
  mutate(GEOID = as.integer(GEOID)) %>% 
  left_join(county_populations, by = c("GEOID" = "fips")) %>% 
  left_join(hospital_data_df, by = c("GEOID" = "FIPS")) %>% 
  mutate(people_per_bed = population / NUM_STAFFED_BEDS)

```

```{r}
quantile(hospital_beds_df$people_per_bed, na.rm = TRUE)

```

```{r}
hospital_beds_df %>% 
  mutate(
    people_per_bed = 
      pmin(7.272759e+03, people_per_bed)
  ) %>% 
  ggplot() +
  geom_sf(aes(fill = people_per_bed), size = 0.01) +
  geom_sf(data = state_boundaries, color = "grey25", fill = NA, size = 0.2) +
  scale_fill_gradientn(
    #limits = c(0, 150),
    #breaks = seq(0, 150, 75),
    colors = blue_pals,
    na.value = MAP_NA_COLOR
  ) +
  guides(
    fill = 
      guide_colorbar(
        barheight = 0.5,
        barwidth = 15,
        title = NULL
      )
  ) +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.5, 0.95),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Number of People Per Hospital Bed in Each County",
    caption = "Definitive Healthcare: USA Hospital Beds  as of June 2020 \nServices7.arcgis.com"
  )
```

Lets load demographic data now

```{r}
demographic_data_df <-
  demographic_data %>% 
  read_excel() %>% 
  mutate(
    STATE = sprintf("%02d",STATE),
    COUNTY = sprintf("%03d", COUNTY),
    STATE = as.character(STATE),
    COUNTY = as.character(COUNTY),
    fips = paste0(STATE, COUNTY),
    fips = as.integer(fips)
  )

```

```{r}
demographic_dataset <-
  county_boundaries %>% 
  mutate(GEOID = as.integer(GEOID)) %>% 
  left_join(demographic_data_df, by = c("GEOID" = "fips"))

```

```{r}
quantile(demographic_dataset$Percent_BAC_Total, na.rm = TRUE)
```

```{r}
demographic_dataset %>% 
  mutate(
    Percent_BAC_Total = 
      pmin(40, Percent_BAC_Total)
  ) %>% 
  ggplot() +
  geom_sf(aes(fill = Percent_BAC_Total), size = 0.01) +
  geom_sf(data = state_boundaries, color = "grey25", fill = NA, size = 0.2) +
  scale_fill_gradientn(
    #limits = c(0, 150),
    #breaks = seq(0, 150, 75),
    colors = green_pal,
    na.value = MAP_NA_COLOR
  ) +
  guides(
    fill = 
      guide_colorbar(
        barheight = 0.5,
        barwidth = 15,
        title = NULL
      )
  ) +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.5, 0.95),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Percentage Black Population by County",
    caption = "Sources: US Census Bureau Race and Ethnicity Distribution in Counties 2018"
  )

```


## Building a model 

First, lets download the data for the area of the counties.

```{r}
counties_area_df <-
  counties_area %>% 
  read_excel() %>% 
  select(
    Areaname,
    fips = STCOU,
    area = LND010190D
  ) %>% 
  filter(Areaname != "UNITED STATES") %>% 
  select(-c(Areaname)) %>% 
  mutate(fips = as.integer(fips))
```

Now, lets combine the datasets into a single dataset so that we have the COVID-19 case and fatality counts for each county, the population, the proportion of households on foodstamps, the proportion of population that is black, and the hospital bed capacity for each county in a single dataframe. 

```{r}
food_benefits_df <-
  income_and_benefits_df %>% 
  select(
    geo_id,
    Percent_total_house_cash_ben,
    Percent_total_house_social_sec_inc,
    Percent_total_house_food_ben
  )
```


```{r}
proportion_black_df <-
  demographic_data_df %>% 
  select(
    fips,
    Percent_BAC_Total
  )
```

Lets add the hospital beds in each county:

```{r}
hospital_beds_per_county <-
  hospital_data_df %>% 
  group_by(FIPS) %>% 
  summarise(beds = sum(NUM_STAFFED_BEDS)) %>% 
  ungroup()
```



```{r}
a <-
  counties_data %>% 
  left_join(counties_area_df, by = c("GEOID" = "fips"))
```

```{r}
b <-
  a %>% 
  left_join(food_benefits_df, by = c("GEOID" = "geo_id"))
```

```{r}
model_1_df <-
  b %>% 
  left_join(proportion_black_df, by = c("GEOID" = "fips"))

```

```{r}
model_1_df <-
  model_1_df %>% 
  mutate(
    population_per_square_mile = population / area,
  )
```

```{r}
model_2_df <-
  model_1_df %>% 
  left_join(hospital_beds_per_county, by = c("GEOID" = "FIPS")) %>% 
  mutate(people_per_hospital_bed = population / beds)
```




Now, lets run a multivariate linear regression model. Our predictor variables are as follows:

* Population per square mile of a county
* Percentage of black population in a county 
* Proportion of Houses that qualify for food benefit
* Number of people per hospital bed 

The outcome variables are number of covid19 cases and number of deaths.

```{r}
model_1 <-
  model_1_df %>% 
  drop_na()
```

```{r}
model_1_a <-
  model_1_df %>% 
  select(
    cases_per_100000,
    population_per_square_mile,
    Percent_BAC_Total,
    Percent_total_house_food_ben
  ) %>% 
  st_set_geometry(NULL) %>% 
  drop_na() %>% 
  filter_all(all_vars(!is.infinite(.)))
```



```{r}
linearmodel <- 
  lm(
    cases_per_100000 ~ population_per_square_mile + Percent_BAC_Total + Percent_total_house_food_ben, data = model_1_a
  )
```

```{r}
summary(linearmodel)
```

```{r}
model_2 <-
  model_2_df %>% 
  drop_na()
```

```{r}
model_2_a <-
  model_2_df %>% 
  select(
    deaths_per_100000,
    population_per_square_mile,
    Percent_BAC_Total,
    Percent_total_house_food_ben,
    people_per_hospital_bed
  ) %>% 
  st_set_geometry(NULL) %>% 
  drop_na() %>% 
  filter_all(all_vars(!is.infinite(.)))

```


```{r}
linearmodel2 <- 
  lm(
    deaths_per_100000 ~ people_per_hospital_bed + population_per_square_mile + Percent_BAC_Total + Percent_total_house_food_ben, data = model_2_a
  )
```

```{r}
summary(linearmodel2)
```

```{r}
linearmodel3 <- 
    lm(
    deaths_per_100000 ~ population_per_square_mile + Percent_BAC_Total , data = model_2
  )

```

```{r}
summary(linearmodel3)
```

## Models at region level.

Lets build a model for states which had the bigger outbreak(NY, NJ, Conneticut, massachusetts, california)

```{r}
model_1_east_coast_df <-
  model_1_df %>% 
  filter(STATEFP %in% east_coast_states)
  
```

```{r}
model_1_east_coast <-
  model_1_east_coast_df %>% 
  select(
    cases_per_100000,
    population_per_square_mile,
    Percent_BAC_Total,
    Percent_total_house_food_ben
  ) %>% 
  st_set_geometry(NULL) %>% 
  drop_na() %>% 
  filter_all(all_vars(!is.infinite(.)))

```

```{r}
linearmodel_east_coast <- 
  lm(
    cases_per_100000 ~ population_per_square_mile + Percent_BAC_Total + Percent_total_house_food_ben, data = model_1_east_coast
  )

```

```{r}
summary(linearmodel_east_coast)

```

```{r}
model_2_east_coast_df <-
  model_2_df %>%
  filter(STATEFP %in% east_coast_states)

```

```{r}
model_2_east_coast <-
  model_2_east_coast_df %>% 
    select(
    deaths_per_100000,
    population_per_square_mile,
    Percent_BAC_Total,
    Percent_total_house_food_ben,
    people_per_hospital_bed
  ) %>% 
  st_set_geometry(NULL) %>% 
  drop_na() %>% 
  filter_all(all_vars(!is.infinite(.)))

```


```{r}
linearmodel2_east_coast <- 
  lm(
    deaths_per_100000 ~ people_per_hospital_bed + population_per_square_mile + Percent_BAC_Total + Percent_total_house_food_ben, data = model_2_east_coast
  )

```

```{r}
summary(linearmodel2_east_coast)
```


Now, lets build models for the southern states

```{r}
model_1_south_df <-
  model_1_df %>% 
  filter(STATEFP %in% south_states)

```

```{r}
model_1_south <-
  model_1_south_df %>% 
    select(
    cases_per_100000,
    population_per_square_mile,
    Percent_BAC_Total,
    Percent_total_house_food_ben
  ) %>% 
  st_set_geometry(NULL) %>% 
  drop_na() %>% 
  filter_all(all_vars(!is.infinite(.)))


```

```{r}
linearmodel_south <- 
  lm(
    cases_per_100000 ~ population_per_square_mile + Percent_BAC_Total + Percent_total_house_food_ben, data = model_1_south
  )

```


```{r}
summary(linearmodel_south)
```


```{r}
model_2_south_df <-
  model_2_df %>%
  filter(STATEFP %in% south_states)

```

```{r}
model_2_south <-
  model_2_south_df %>% 
    select(
    deaths_per_100000,
    population_per_square_mile,
    Percent_BAC_Total,
    Percent_total_house_food_ben,
    people_per_hospital_bed
  ) %>% 
  st_set_geometry(NULL) %>% 
  drop_na() %>% 
  filter_all(all_vars(!is.infinite(.)))


```


```{r}
linearmodel2_south <- 
  lm(
    deaths_per_100000 ~ people_per_hospital_bed + population_per_square_mile + Percent_BAC_Total + Percent_total_house_food_ben, data = model_2_south
  )

```

```{r}
summary(linearmodel2_south)
```

## Map of Population Density 

```{r}
pop_density_df <-
  a %>% 
  mutate(pop_density = population / area)
```

```{r}
quantile(pop_density_df$pop_density, na.rm = TRUE)
```


```{r}
pop_density_df %>% 
  mutate(pop_density = pmin(111.301638, pop_density)) %>% 
  ggplot() +
  geom_sf(aes(fill = pop_density), size = 0.01) +
  geom_sf(data = state_boundaries, color = "grey25", fill = NA, size = 0.2) +
  scale_fill_gradientn(
    #limits = c(0, 150),
    #breaks = seq(0, 150, 75),
    labels = labels,
    colors = yellow_pal,
    na.value = MAP_NA_COLOR
  ) +
  guides(
    fill = 
      guide_colorbar(
        barheight = 0.5,
        barwidth = 15,
        title = NULL
      )
  ) +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.5, 0.95),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  labs(
    title = "US Population Density(by Counties)",
    caption = "Source: US Census Bureau Population Distribution in Counties 2018, \nUS Census Bureau County Land Areas 2011"
  )
```

